Contracts for unions and intersections are hard for Nickel

### 2021-06-07 paper.tex
Remove trailing spaces
Rework of section 4 (subsection on intersections)

Static typing is about enforcing that some property about a (potentially
unknown) value always holds, no matter the execution path. A dynamic check, on
the other hand, just need to verify a property for a specific and known value,
which is usually much easier. For example, dependent function types are complex
and mostly implemented in proof assistants (although they are slowly making
their way in some mainstream languages), whereas the corresponding dependent
function contracts can be implemented in a few line of code (they are
implemented in e.g. Racket~\cite{}).

The gist of the present paper is to bring evidence that in the case of union and
intersection, the dynamic checking part is complex as well: \emph{unions and
intersections are fundamentally hard}.

@@

In Nickel, the failure of a function contract can always be traced back to a
single call. For example, take the function \nickel{f} with a simple contract
attached of Figure~\ref{fig:pos-to-pos}. The program fails with a contract error
blaming \nickel{f}. Indeed, the second call \nickel{f 5} fails because the
return value violates the \nickel{Positive} contract. The first call to
\nickel{f} does not matter, and \nickel{f 5} is a single and independent witness
of the violation of the contract. The user can be pointed to one location in
practice.

This single witness property can be justified as follows. Apart from the error
reporting part (although this is the crucial bit in practice!), the current
contract system of Nickel can be implemented purely as a library, requiring only
a \nickel{fail} primitive to abort the execution. In practice, applying a
function contract to \nickel{f} replaces it with a \nickel{f'} that additionally
performs the required checks. Thus, since the core language is pure (the
presence of \nickel{fail} isn't relevant here), the failure of \nickel{f' 5}
must be independent of its environment and of any previous call to \nickel{f'}.

@@

This program must fail, because \nickel{f 10} is a witness of \nickel{f} failing
the contract \nickel{Positive -> NonPositive}, and \nickel{f 5} is a witness of
\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
Figure~\ref{fig:pos-to-pos}, removing only one the call makes the program
succeed! Indeed, the first call only fails the \nickel{Positive -> NonPositive}
part of the union, while the second one only fails the \nickel{Positive ->
NonPositive} part. In this particular example, there can't be a single call that
is a witness of the violation of the contract: a minimum of two are needed.

This behavior indicates that union contracts introduce side-effects. The result
of \nickel{f 5} now depends on the previous execution and more generally on any
prior call to \nickel{f}. This behavior of union contracts break the
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}. The
candidate example of Figure~\ref{fig:optimizations-let-floating-ex} can't be
optimized in general.

Figure~\ref{fig:optimized-programs} illustrates this point further. There is an
original program, and an optimized version the common subexpression \nickel{f 1}
has been eliminated. While equivalent in a pure language without contracts,
these two programs behave differently here:

@@

In the original version, each partial application \nickel{f 1} gives rise to a
fresh instance of the contract \nickel{Bool -> Num @| Bool -> Str}. These
instances are independent, and can pick a different component of the union to
satisfy. Although \nickel{f} doesn't actually respect the contract, these calls
are not enough to prove so. In the optimized version, \nickel{g} is endowed with
a single contract, that must pick one of the two components of the union. There,
the two calls refers to the same union contracts, and shows that \nickel{f}
violate its initial contract.

@@

To sum up, the addition of union contracts introduce side-effects in a pure
language. Side-effects have well-known pitfalls: 
\begin{itemize}
    \item For the programmer, they are hard to reason about. They prevent local
        reasoning. In our previous examples, removing or adding a function call
        somewhere can toggle a failure in a call at a totally different
        location.
    \item For the interpreter (or compiler), they inhibit many optimizations and
        program transformations.
\end{itemize}

@@

\subsection*{Intersection of user-defined contracts}
\label{sec:flat-and-inter}

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially. This intuition works for simple contracts: surely
checking that \nickel{x | Natural @& Odd} amounts to check that \nickel{x |
Natural} and \nickel{x | Odd}. Unfortunately, this doesn't scale to higher-order
contracts.

@@

works for a lot of simple contracts: Stop for
a second and think how you could implement an intersection contract. The first
thought you may have is "just apply both contracts", as shown on the
Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if it
were that simple we wouldn't be writing this paper.  This idea starts to show
problems when you first combine it with higher order contracts, since it would
mean not complying with at least one of these contracts negatively would entail
breaking the whole thing, as shown on the second example on the same Figure,
that would fail.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x | A @& B (*@ $\simeq$ @*) (x | A) | B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Problem]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{Implementations of intersection contracts: first iteration}
\label{fig:inter-contracts}
\end{figure}

Not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the shared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
something like \nickel{1 | Str}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Num -> Num}) has
also tried to raise negative blame on the same parameter ()\nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
\end{lstlisting}
\caption{Implementations of intersection contracts: second iteration}
\label{fig:inter-contracts-2}
\end{figure}

On the other hand, if you were to do something like
\nickel{g true}, that would fail also negatively the contract
\nickel{Str -> Str} and, when checking if \nickel{true}
is a correct parameter for \nickel{Num -> Num} it would check
if the other side of the intersection has failed, and since it
has, a violation would be raised and the whole program would
correctly fail.

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with user defined contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check the contract \nickel{C}, by
checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g | (String -> String) @& C
      = (fun x => x) 
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
\nickel{g} is never applied to \nickel{0}, in fact, it is never applied
to anything.
Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
should not have been used inside the evaluation of \nickel{C}, which
adds an extra layer of complexity: contracts should be guarded, and execution
depends on the shape of the surrounding evaluation context.

>>>>

Static typing is about enforcing that some property about a (potentially
unknown) value always holds, no matter the execution path. A dynamic check, on
the other hand, just needs to verify a property for a specific and known value,
which is usually much easier. For example, dependent function types are complex
and mostly implemented in proof assistants (although they are slowly making
their way in some mainstream languages), whereas the corresponding dependent
function contracts can be implemented in a straightforward way (they are
supported in e.g. Racket~\cite{}).

The gist of the present paper is to bring evidence that in the case of unions
and intersections, the dynamic checking part is complex as well: \emph{unions
and intersections are fundamentally hard}.

@@

In Nickel, the failure of a function contract can always be traced back to a
single call. For example, take the function \nickel{f} with a simple contract
attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
error blaming \nickel{f} because the return value of the second call \nickel{f 5}
violates the \nickel{Positive} contract. The first call to \nickel{f} does
not matter, and \nickel{f 5} is a single and independent witness of the contract
violation. The user is pointed to this one location in practice.

This single witness property can be justified as follows. Apart from the error
reporting part (although this is the crucial bit in practice!), the current
contract system of Nickel can be implemented purely as a library, requiring only
a \nickel{fail} primitive to abort the execution. In practice, applying a
function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
additional checks. Thus, since the core language is pure (forgetting the
presence of \nickel{fail}, which isn't relevant), the failure of \nickel{f' 5}
must be independent of its environment and of any previous call to \nickel{f'}.

@@

This program must fail, because \nickel{f 10} is a witness of \nickel{f} failing
the contract \nickel{Positive -> NonPositive}, and \nickel{f 5} is a witness of
\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to the
example from Figure~\ref{fig:pos-to-pos}, removing only one of the calls makes the
program succeed! Indeed, each call only unveils the violation of one component
of the union. In this example, a single call to \nickel{f} that would be the
witness of the violation of the whole contract doesn't even exist: a minimum of
two are always needed.

This behavior indicates that union contracts introduce side-effects. The result
of \nickel{f 5} now depends on the previous execution and more specifically on
any prior call to \nickel{f}. This behavior of union contracts break the
property~\ref{eq:cbn-expansion} introduced in Section~\ref{sec:optimizations},
that is required to perform CSE-like optimizations in all generality. The
candidate example of Figure~\ref{fig:optimizations-let-floating-ex} in
Section~\ref{sec:optimizations} can't be optimized in general.

Figure~\ref{fig:optimized-programs} illustrates this point further. It contains
an original program and an optimized version where the common subexpression
\nickel{f 1} has been eliminated. While equivalent in a pure language without
contracts, these two programs behave differently because of unions:

@@

In the original version, each partial application \nickel{f 1} gives rise to a
fresh instance of the contract \nickel{Bool -> Num @| Bool -> Str}. These
instances are independent, and can pick a different component of the union to
satisfy. Although \nickel{f} doesn't actually respect the contract, these calls
are not enough to prove so. In the optimized version, \nickel{g} is endowed with
a single contract, that must pick one of the two components of the union. There,
the two calls refer to the same union contract, and shows that \nickel{f} does
violate its initial contract.

@@

To sum up, the addition of union contracts introduce side-effects in a pure
language. Side-effects have well-known pitfalls:
\begin{itemize}
    \item For the programmer, they are hard to reason about. They prevent local
        reasoning. In our previous examples, removing or adding a function call
        somewhere can toggle a failure in a call at a totally different
        location.
    \item For the interpreter (or compiler), side-effects inhibit many optimizations and
        program transformations.
\end{itemize}

@@

\subsection*{Intersection with user-defined contracts}
\label{sec:flat-and-inter}

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially:

\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
decomposition}]
M | A @& B (*@ $\simeq$ @*) (M | A) | B
\end{lstlisting}

This intuition works for simple contracts: checking that \nickel{x | Natural @&
Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
Unfortunately, this doesn't scale to higher-order contracts.
Figure~\ref{fig:inter-contracts} illustrates the use of an intersection to model
a simple overloading of the identity function. If we were to apply the naive
equivalence above, the argument \nickel{1} would fail the \nickel{Str -> Str}
contract. The naive decomposition actually implies

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
\end{lstlisting}

This isn't the correct semantics for overloading. With this implementation, our
contract for identity from Figure~\ref{fig:inter-contracts} would always
fail because no argument satisfies \nickel{Num @& Str}.

% Stop for a second and think how you could implement an intersection contract.
% The first thought you may have is "just apply both contracts", as shown on the
% Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if
% it were that simple we wouldn't be writing this paper.  This idea starts to
% show problems when you first combine it with higher order contracts, since it
% would mean not complying with at least one of these contracts negatively would
% entail breaking the whole thing, as shown on the second example on the same
% Figure, that would fail.

\begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Implementation]
% x | A @& B (*@ $\simeq$ @*) (x | A) | B
% \end{lstlisting}
\begin{lstlisting}[language=nickel]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{In}
\label{fig:inter-contracts}
\end{figure}

In an higher-order intersection contract, blame is raised when:
\begin{description}
    \item[Negative blame] The argument fails \emph{both} components.
    \item[Positive blame] The function fails \emph{one} component that the
        argument previously satisfied.
\end{description}

% Not all hope is lost, we could pass some extra shared state
% to each one of the sub contracts and let them only fail if the other
% side of the contract has failed as well, as shown in Figure
% \ref{fig:inter-contracts-2}, where the shared state is
% represented by a label \nickel{l}.
% This idea is still quite reasonable, for instance, when executing
% \nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
% that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
% something like \nickel{1 | Str}.
% Of course, this fails, but before raising a violation, this contract could
% check if the other side of the intersection (\nickel{Num -> Num}) has
% also tried to raise negative blame on the same parameter ()\nickel{1}), since
% it hasn't, the contract would return normally, and would mark itself as failed.

% On the other hand, if you were to do something like
% \nickel{g true}, that would fail also negatively the contract
% \nickel{Str -> Str} and, when checking if \nickel{true}
% is a correct parameter for \nickel{Num -> Num} it would check
% if the other side of the intersection has failed, and since it
% has, a violation would be raised and the whole program would
% correctly fail.
%
% This idea, which is an oversimplification of the ideas
% presented by Williams, Morris, and Wadler, seems to be spot on.

To fix the naive implementation, the interpreter can share state between the
sub-contracts, in order to decide if blame must be raised or not when a
sub-contract fails:

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
\end{lstlisting}

Shared state is represented by the label \nickel{l}. This is in essence the
approach proposed by Williams, Morris, and Wadler in~\cite{RootCauseOfBlame}.

% \begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Implementation]
% x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
% \end{lstlisting}
% \caption{Implementations of intersection contracts: second iteration}
% \label{fig:inter-contracts-2}
% \end{figure}

% However, it presents a major problem, it's not compatible with user defined contracts,
% as presented in Section \ref{sec:flat-contracts}.
% Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
% implementation discussed previously were to be used, we may end up with
% something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
% \nickel{l} represents the shared state.
% At this point, the intuitive thing to do is to check the contract \nickel{C}, by
% checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
% since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
% fails negatively.

However, their approach has a major drawback: it is not straightforwardly
compatible with user-defined contracts (introduced in Section
\ref{sec:flat-contracts}). The issue is similar to our initial issue with higher-order
contracts and the naive decomposition: user-defined contracts may apply
functions and thus make a sub-contract of the intersection fail, but this
failure shouldn't always result in raising blame. An example is given on Figure
\ref{fig:inter-flat-contracts}. Decomposing using the shared state approach, we
end up with:

\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Stateful
decomposition}]
((fun x => x) | (Str -> Str)[l]) | C[l]
\end{lstlisting}

where \nickel{l} represents the shared state.  At this point,
applying the \nickel{C} contract results in evaluating:

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
((fun x => x) | (Str -> Str)[l]) 0 == 0.
\end{lstlisting}

Applying a function wrapped in a \nickel{Str -> Str} contract to \nickel{0}
fails negatively. This is not the expected behavior, since the identity function
does respect semantically both contracts. As opposed to built-in higher-order
contracts, user-defined contracts are black-box from the interpreter's point of
view, and it is thus not obvious how to extend the shared state approach to
handle user-defined contracts. The authors of~\cite{RootCauseOfBlame}
acknowledge this fact: they decide to simply drop user-defined contracts
altogether.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => f 0 == 0 in
let g | (String -> String) @& C
      = fun x => x
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

Once again, intersection contracts introduce side-effects in the picture. What's
more, these side-effects interact with user-defined contracts in a non-trivial
way, while they are an important feature for validation.
\unsure{(Yann) Should we talk about thiemann here, who do support user-defined
contracts?}
% But this should not have happened, since in the program of Figure
% \ref{fig:inter-flat-contracts} \nickel{g} is never applied to \nickel{0}, in
% fact, it is never applied to anything.  Checking if \nickel{C} has already
% failed, and marking \nickel{String -> String} as failed negatively on \nickel{l}
% is not enough, the contract \nickel{String -> String} should not have been used
% inside the evaluation of \nickel{C}, which adds an extra layer of complexity:
% contracts should be guarded, and execution depends on the shape of the
% surrounding evaluation context.

@@
